From physical principles to ???

### 2016-05-13 2015CompleteDerivation/completeDerivation.tex
More review on classical hamiltonian systems

On a state space with discrete topology and a single state variable we simply have $\Lambda_U (\mathcal{c}) = \sum \limits_{q=a}^b \rho(q)$. For states identified by a single continuous state variable we would expect $\Lambda_U (\mathcal{c}) = \int_a^b \rho (q) dq$. By a simple change of state variable, we see that this does not work in general: $\rho(q')= \rho(q) dq/dq'$. This would make $\rho$ a function not just of the state $\mathcal{s}$, but also of the state variables we are using. This is inconsistent with our previous definition. Moreover if the transformation is not differential, the density is not even well-defined. This tells us that $\mathcal{S}$ cannot be any manifold: it has to be one that allows state-variable-independent densities. Under arbitrary change of state variables, the density has to remain the same, which means that the Jacobian of the transformation must exist and must be unitary.

For the Jacobian to exist $\mathcal{S}$ has to be a differentiable manifold; that is the transformation between state variables is always differentiable. This way, at least, the density is well defined for any choice of state variables.

If $\rho$ is to remain unchanged and the Jacobian to be unitary we cannot simply change one state variable as we please. If we change one at least another has to change in some coordinated way such that the Jacobian of the total transformation is unitary. This means that we cannot change physical units of all the variables as we please: they are part of a unit system. For example, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. As these are states for infinitesimal parts, it will be a local object in $\mathcal{Q}$. As we saw before, it cannot be a point as the density $\rho(q')= \rho(q) dq/dq'$ is not invariant. It cannot be a vector $v(q) e_q$, as the density $\rho(q', v')= \rho(q, v) (dq/dq')^2$ is not invariant. A covector $k(q) dq$ is instead fine as $\rho(q', k')= \rho(q, k) (dq/dq') (dq'/dq) = \rho(q, k)$. In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

@@

	If we assume the whole system is deterministic and reversible but each part is not, then their evolution can depend on the states of the other parts. But this does not solve the problem entirely. Let's ignore the question of how we were able to define the states of the parts. Let's also ignore that the type of forces one would typically use (i.e. ones that conserve energy or ones with scalar and vector potentials) are precisely the ones we derive by assuming the pieces are not interacting with each other. The problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if the particles are in principle interacting, the system is described by degrees of freedom that evolve independently, oblivious to each other.

	The same type of problem arises on the whole system when we consider its interaction with our probe or other parts of the environment. Say $a \in A$ is the state of our system and $b \in B$ the state of our probe, the future state $\hat{a}=f(a)$ because the system is deterministic, but $\hat{b}=g(a,b)$ because the probe needs to be affected by the system. This means $(a, b) \mapsto (\hat{a}, \hat{b})$ is not a reversible map. In other words: we can't expect to have deterministic and reversible evolution at all levels of decomposable systems where the parts interact in any physically meaningful way.

>>>>

On a state space with discrete topology and a single state variable we simply have $\Lambda_U (\mathcal{c}) = \sum \limits_{q=a}^b \rho_\mathcal{c}(q)$. For states identified by a single continuous state variable we would expect $\Lambda_U (\mathcal{c}) = \int_a^b \rho_\mathcal{c} (q) dq$. By a simple change of state variable, we see that this does not work in general: $\rho_\mathcal{c}(q')= \rho(q)_\mathcal{c} dq/dq'$. This would make $\rho_\mathcal{c}$ a function not just of the state $\mathcal{s}$, but also of the state variables we are using. This is inconsistent with our previous definition. Moreover if the transformation is not differential, the density is not even well-defined. This tells us that $\mathcal{S}$ cannot be any manifold: it has to be one that allows state-variable-independent densities. Under arbitrary change of state variables, the density has to remain the same, which means that the Jacobian of the transformation must exist and must be unitary.

For the Jacobian to exist the transformation between state variables has to be differentiable. Therefore a manifold that guarantees densities to be well defined is one that guarantees that state variable transformations are differentiable: a differentiable manifold. $\mathcal{S}$ has a differentiable structure in the sense that it is on that structure that densities are defined. In the same way that discontinuous changes of state variable can be detected because they are incompatible with the topology, non differential changes of state variable can be detected because they are incompatible with state-variable-independent densities.

If $\rho_\mathcal{c}$ is to remain unchanged and the Jacobian to be unitary we cannot simply change one state variable as we please. If we change one at least another has to change in some coordinated way such that the Jacobian of the total transformation is unitary. This means that we cannot change physical units of all the variables as we please: they are part of a unit system. For example, changing units of position will also change units of momentum. Let $\mathcal{Q}$ be the manifold charted by the minimum set of state variables that define such system of units. An element of $\mathcal{S}$ will be a geometrical object in such space. For example, it will be a point, a distance, two points, the ratio between two distances, the inverse of a distance: something that is expressible in the unit defined. As the particle state represent configurations of infinitesimal parts, it will be a local object in $\mathcal{Q}$, defined around a point. As we saw before, it cannot be a point itself as the density $\rho(q')_\mathcal{c}= \rho(q)_\mathcal{c} dq/dq'$ is not invariant. It cannot be a vector $v(q) e_q$, as the density $\rho_\mathcal{c}(q', v')= \rho(q, v)_\mathcal{c} (dq/dq')^2$ is not invariant. A covector $k(q) dq$ is instead fine as $\rho_\mathcal{c}(q', k')= \rho_\mathcal{c}(q, k) (dq/dq') (dq'/dq) = \rho_\mathcal{c}(q, k)$. In other words: a covector is the simplest local geometrical object that allows coordinate independent densities. We can identify $\mathcal{S}$ as the cotangent bundle $T^*\mathcal{Q}$, the space of all possible covectors.

@@

	If we just assume the whole system is deterministic and reversible, then each part can evolve depending the states of the other parts. But this does not solve the problem entirely. Let's ignore the question of how we were able to define the states of the parts. Let's also ignore that the type of forces one would typically use (i.e. ones that conserve energy or ones with scalar and vector potentials) are precisely the ones we derive by assuming the pieces are not interacting with each other. The problem is that we can always locally separate the evolution into independent degrees of freedom. For example, the position and momentum of two particles may affect each other during the evolution, but the average and difference in position and momentum may evolve independently.\footnote{Also note that any Hamiltonian is locally isomorphic to a free particle (citation needed)} We can always find such local decomposition, and the easiest way to see that is in terms of the transported variables: they retain the original value, they clearly evolve independently. And since the pieces are infinitesimal, a local decomposition is all that is needed. While such description may be cumbersome to achieve in practice, conceptually it is still possible. So, even if the particles are in principle interacting, the system is described by degrees of freedom that evolve independently, oblivious to each other.

	The same type of problem arises on the whole system when we consider its interaction with our probe or other parts of the environment. Say $a \in A$ is the state of our system and $b \in B$ the state of our probe, the future state $\hat{a}=f(a)$ because the system is deterministic, but $\hat{b}=g(a,b)$ because the probe needs to be affected by the system. This means $(a, b) \mapsto (\hat{a}, \hat{b})$ is not a reversible map. In other words: we can't expect to have deterministic and reversible evolution at all levels of decomposable systems if the parts interact in any physically meaningful way.
